ISYE 6501

### 2022-09-28 hw5-1/submission.Rmd
submission file created

---
title: "Random Forest | Ques 10.1"
author: "Ashish Dhiman"
date: "2022-09-26"
output: html_document
---

### Read Data

```{r, warning=FALSE}
set.seed(77)
library(randomForest)
library(ggplot2)

crime_df = read.table("../uscrime.txt", sep="\t", header= TRUE)
dim(crime_df)
```

#### Data and variable meaning:

Criminologists are interested in the effect of punishment regimes on crime rates. This has been studied using aggregate data on 47 states of the USA for 1960.

| Variable | Description                                                                            |
|--------------------|----------------------------------------------------|
| M        | percentage of males aged 14--24 in total state population                              |
| So       | indicator variable for a southern state                                                |
| Ed       | mean years of schooling of the population aged 25 years or over                        |
| Po1      | per capita expenditure on police protection in 1960                                    |
| Po2      | per capita expenditure on police protection in 1959                                    |
| LF       | labour force participation rate of civilian urban males in the age-group 14-24         |
| M.F      | number of males per 100 females                                                        |
| Pop      | state population in 1960 in hundred thousands                                          |
| NW       | percentage of nonwhites in the population                                              |
| U1       | unemployment rate of urban males 14--24                                                |
| U2       | unemployment rate of urban males 35--39                                                |
| Wealth   | wealth: median value of transferable assets or family income                           |
| Ineq     | income inequality: percentage of families earning below half the median income         |
| Prob     | probability of imprisonment: ratio of number of commitments to number of offenses      |
| Time     | average time in months served by offenders in state prisons before their first release |
| Crime    | crime rate: number of offenses per 100,000 population in 1960                          |

```{r}
crime_df$So <- factor(crime_df$So)
```

@@

@@

@@

We have created uniform CV folds here

@@

#### Impact of ntree

@@

As

>>>>

---
title: "ISYE 6501"
output:
  html_document:
    df_print: paged
  pdf_document: default
date: "2022-09-04"
---

# Submission HW4 \| Fall 22

-   Ashish Dhiman \| [ashish.dhiman\@gatech.edu](mailto:ashish.dhiman@gatech.edu){.email}

-   Abhinav Arun \| [aarun60\@gatech.edu](mailto:aarun60gatech.edu){.email}

-   Anshit Verma \| [averma373\@gatech.edu](mailto:averma373@gatech.edu){.email}

[**\*Analysis Notes are marked with Red header: #Analysis**]{style="color:red"}

```{r, warning=FALSE, message=FALSE}
set.seed(77)

library(randomForest)
library(ggplot2)
library(cowplot)
library(ggplot2)
library(reshape2)
library(grid)
library(MASS)
library(rpart)
library(rpart.plot)
```

```{r}
crime_data<-read.table(file = "./uscrime.txt", sep = "\t",header=TRUE)
head(crime_data)
dim(crime_data)
summary(crime_data)
```

## EDA

### Data Distribution and Pairwise Corelation

```{r}
my_plots <- lapply(names(crime_data), function(var_x){
  p <- 
    ggplot(crime_data) +
    aes_string(var_x)

  if(var_x %in% list("So")) {
    p <- p + geom_bar()

  } else {
    p <- p + geom_density()
  }

})

plot_grid(plotlist = my_plots)

### Correlation
cormat <- round(cor(crime_data),2)
cormat[upper.tri(cormat)] <- NA
melted_cormat <- melt(cormat)
# plotting the correlation heatmap

ggplot(data = melted_cormat, aes(x=Var1, y=Var2,
                                   fill=value)) +
geom_tile() +
geom_text(aes(Var1, Var2, label = value),
          color = "white", size = 3)
```

## Decision Tree

#### Training and evaluating on the same training dataset (not a good method for generalizing) , better to do cross validation

```{r}
# Vanilla decision tree with a fixed value of complexity parameter of 0.01 and minsplit of 20
reg_tree<-rpart(Crime~.,data=crime_data,method="anova",control=c(cp=0.01,minsplit=20))
rmse<-sqrt(sum((crime_data$Crime - predict(reg_tree,type=c("vector")))**2)/nrow(crime_data))
rmse
```

```{r}
rpart.plot(reg_tree)
```

[**\*Analysis Notes are marked with Red header: #Analysis**]{style="color:red"}

-   Observations & Points for further exploration:

-   The above decision tree uses 3 variables NW , Pop, LF and Po1 to create a decision tree. (2 of them Pop and Po1 have a linear correlation with dependent variable i.e. Crime)

-   One food for thought is that this is a regression tree and we are imputing the mean values of all the points that fall in a leaf or terminal node , so incase if we have an outlier , then those will drive the value of mean and hence cause higher RMSE or SSE. So , it would be good to see the effect of removing or capping outliers and then performing decision tree.

-   The smaller the value of complex parameter , the model is expected to have high variance as a split would be performed if we get such a marginal increase in entropy . Therefore , to prevent our model from overfitting amd having high variance , it is better if we not choose too small values of cp i.e. complexity parameter.

-   The default value of max depth parameter here is 30 but given that we have only 47 nodes in our dataset , it dfoes not make much sense to have tree models greater than or equal to 3 or 4 levels of depth

-   minsplit hyperparameter : The default value of this parameter is 20 i.e. a split would be performed if we have atleast 20 datapoints in a node , although it is generally good to have higher values of "minsplit" parameter to avoid overfitting but since in this problem statement , the total \# of datapoints is only 47 , so this is one parameter where we can tweak in a bit .

-   minbucket hyperparameter : Similarly , this hyperparameter means the number of datpoints in a terminal node and thus here also , it makes sense to have a smaller value for this hyperparameter as we do not have large dataset for training our model.

#### Perform hyperparameter tuning minsplit and cp(complexity hyperparameter)

```{r}
folds <- sample(rep(1:5, length.out = nrow(crime_data)), size = nrow(crime_data), replace = F)
pred_table<-data.frame(min_split=numeric(0),cp=double(0),rmse=double(0))
for(j in seq(0.001,0.02,0.001)){
for(i in seq(5,25,2)){
  CV_mse <- lapply(1:5, function(x){ #5 corresponds to the number of folds defined earlier
  model <- rpart(Crime ~ ., data = crime_data[folds != x,],method="anova",control=c(minsplit=i,cp=j))
  preds <- predict(model,  crime_data[folds == x,], type="vector")
  sum_sse<-sum((crime_data[folds==x,"Crime"]-preds)**2)
  rmse<-sqrt(sum_sse/nrow(crime_data[folds==x,]))
  #total_rmse<-total_rmse+rmse
  return(rmse)
   })
pred_table[nrow(pred_table)+1,]=c(i,j,mean(unlist(CV_mse)))
}
}
```

# Identifying the optimal value of the hyperparameters cp and min_split

```{r}
options(scipen=999)
pred_table[which.min(pred_table$rmse),c("min_split","cp")]
```

-   From the above output , we get the the optimal value of cp as 0.001 and min_split as 15

```{r}
reg_cv_tree<-rpart(Crime~.,data=crime_data,method="anova",control=c(cp=0.001,minsplit=15))
rpart.plot(reg_cv_tree)
```

-   [**\*Analysis Notes are marked with Red header: #Analysis**]{style="color:red"}
-   Note: The above set of optimal parameters reproduces the same decision tree that we had done initially with fixed values of these 2 parameters.This can be because of less \# of datapoints. We can try reducing the number of minsplits which will lead to a higher variance model and we can try iterating to get the optimal maxDepth hyperparameter.

```{r}
pred_table<-data.frame(min_split=numeric(0),maxDepth=double(0),rmse=double(0))
for(j in seq(2,15,1)){
for(i in seq(5,25,2)){
  CV_mse <- lapply(1:5, function(x){ #5 corresponds to the number of folds defined earlier
  model <- rpart(Crime ~ ., data = crime_data[folds != x,],method="anova",control=c(minsplit=i,maxdepth=j))
  preds <- predict(model,  crime_data[folds == x,], type="vector")
  sum_sse<-sum((crime_data[folds==x,"Crime"]-preds)**2)
  rmse<-sqrt(sum_sse/nrow(crime_data[folds==x,]))
  #total_rmse<-total_rmse+rmse
  return(rmse)
   })
pred_table[nrow(pred_table)+1,]=c(i,j,mean(unlist(CV_mse)))
}
}  
#rpart.plot(rpart(Crime ~ ., data = crime_data,method="anova",minsplit=10,cp=0.01))
#plot(pred_table)
```

-   [**\*Analysis Notes are marked with Red header: #Analysis**]{style="color:red"}
-   Identifying the optimal value of the hyperparameter maxDepth (which comes out to be around 2 with min_split within range of 5-15 (rmse is same for different values of min_split))

```{r}
options(scipen=999)
pred_table[which.min(pred_table$rmse),c("maxDepth")]
```

```{r}
reg_tree_final<-rpart(Crime~.,data=crime_data,method="anova",control=c(cp=0.001,minsplit=15,maxdepth=2))
rmse_final<-sqrt(sum((crime_data$Crime - predict(reg_tree_final,type=c("vector")))**2)/nrow(crime_data))
rmse_final
```

**Conclusion : The RMSE value for the decision tree model using the optimal values of the hyperparameter is `r round(rmse_final,2)`**

## Random Forest

```{r}
crime_df <- crime_data
crime_df$So <- factor(crime_df$So)
```

@@

[**\*Analysis Notes are marked with Red header: #Analysis**]{style="color:red"}

@@

Variable Importance here is given basis two metrics, drop in MSE when the variable is not included in feature set, and second the Information gain (or purity) of a node, if the variable is used to split it.

@@

[**\*Analysis Notes are marked with Red header: #Analysis**]{style="color:red"}

We have created uniform CV folds here.

@@

@@
